## Task

1.  Do 5 Code Wars problems in R. Share your solutions here:

    ```{r}

    # Problem 1 I solved in R:

    #The two oldest ages function/method needs to be completed. It should take an array of numbers as its argument and return the two highest numbers within the array. The returned value should be an array in the format [second oldest age,  oldest age].

    #The order of the numbers passed in could be any order. The array will always include at least 2 items. If there are two or more oldest age, then return both of them in array format.

    #For example (Input --> Output):

    [1, 2, 10, 8] --> [8, 10]
    [1, 5, 87, 45, 8, 8] --> [45, 87]
    [1, 3, 10, 0]) --> [3, 10]                          
    })

    #solution
      
     two_oldest_ages <- function(ages) {
      return(sort(ages, decreasing = TRUE)[1:2])
    }

     ---------------------------------------------------------------------
       
     # Problem 2 I solved in R:
     
    #We need a function that can transform a string into a number. What ways of achieving this do you know?


    #Examples
    "1234" --> 1234
    "605"  --> 605
    "1405" --> 1405
    "-7" --> -7

    # solution

    string_to_number <- function(str) {
      return(as.numeric(str))
    }

    ---------------------------------------------------------------------
      
      # Problem 3 I solved in R:
      
      #In this simple assignment you are given a number and have to make it negative. But maybe the number is already negative?

    #Examples
    make_negative(1);  # return -1
    make_negative(-5); # return -5
    make_negative(0);  # return 0

    # solution

    make_negative <- function(num) {
      if (num >= 0) {
        return(num * -1)
      } else {
        return(num)
      }
    }
    ----------------------------------------------------------------------
      
      # Problem 4 I solved in R : Write a function that takes two arguments, a string of lowercase letters and a positive integer, and returns a new string with every kth letter removed.
      
      #solution
      
      remove_every_kth_letter <- function(str, k){
      letters <- strsplit(str, "")[[1]]
      new_str <- ""
      for (i in 1:length(letters)){
        if (i %% k != 0){
          new_str <- paste0(new_str, letters[i])
        }
      }
      return(new_str)
    }
    ----------------------------------------------------------------------
      #Problem 5 I solved in R : Write a function that takes a string of words and returns the length of the shortest word.
      
      #solution
      
      shortest_word <- function(str){
      words <- strsplit(str, " ")[[1]]
      shortest <- nchar(words[1])
      for (i in 2:length(words)){
        if (nchar(words[i]) < shortest){
          shortest <- nchar(words[i])
        }
      }
      return(shortest)
    }

    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}

    fizz_buzz <- function(num) {
      if (num %% 3 == 0 && num %% 5 == 0) {
        return("fizzbuzz")
      } else if (num %% 3 == 0) {
        return("fizz")
      } else if (num %% 5 == 0) {
        return("buzz")
      } else {
        return(num)
      }
    }

    fizz_buzz(9)


    fizz_buzz(20)


    fizz_buzz(45)


    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called Mario Stairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.


    marioStairs <- function(n) {
      for (i in 1:n) {
        cat(paste0(rep("X", i), collapse = ""), "\n")
      }
    }


    marioStairs(100)

    ```
